---
title: "HCR-FISH-data-analysis"
author: "Tajda Klobučar"
date: "2025-02-10"
output: html_document
---

#### Libraries

```{r include=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(superb))
suppressPackageStartupMessages(library(readxl))

```
## Imported all quantifications (number of foci, their size and intensity) from .csv obtained from FIJI analysis. One file was generated per well imaged (A1, A2_0 to A2_6), all of them containing probes against three different transcripts labeled with 3 different fluorophores (488, 594 or 647). These files were generated using macros and raw image files available on Zenodo (10.5281/zenodo.13860869).

## First merge them by fluorophores and group - smOOPS or non-smOOPs (also reffered here as control).
## Calculate Total intensity values (mean*area).

## Load the data. 

```{r include=FALSE}
WellA1_488 <- read_excel("WellA1_488.xlsx")
WellA1_640 <- read_excel("WellA1_640.xlsx")
WellA1_561 <- read_excel("wellA1_A1_561.xlsx")
WellA2_0_488 <- read_excel("WellA2_0_488.xlsx")
WellA2_0_640 <- read_excel("WellA2_0_640.xlsx")
WellA2_5_488 <- read_excel("WellA2_5_488.xlsx")
WellA2_6_488 <- read_excel("WellA2_6_488.xlsx")
WellA2_0_561 <- read_excel("wellA2_0_561.xlsx")
WellA2_5_561 <- read_excel("wellA2_5_561.xlsx")
WellA2_6_561 <- read_excel("wellA2_6_561.xlsx")
WellA2_5_640 <- read_excel("WellA2_5_640.xlsx")
WellA2_6_640 <- read_excel("WellA2_6_640.xlsx")

WellA2_3_488 <- read_excel("WellA2_3_488.xlsx")
WellA2_4_488 <- read_excel("WellA2_4_488.xlsx")
WellA2_3_561 <- read_excel("wellA2_3_561.xlsx")
WellA2_4_561 <- read_excel("wellA2_4_561.xlsx")
WellA2_4_640 <- read_excel("WellA2_4_640.xlsx")

# And load data with gene_name and gene_id with expression values also. 
targets_info <- read_excel("C:/Users/tajdak/UK Dementia Research Institute Dropbox/Tajda Klobucar_DRI/KI/Protocols/Semi_oops project/HCR_FISH design/IJS_imaging/july_2024/Particle_analysis/targets_info.xlsx")

```

## First merge them by fluorophores and group - smOOPS or non-smOOPs (also reffered here as control).
## Calculate Total intensity values (mean*area).

```{r}
smoops_all_green <- rbind(WellA1_488, WellA2_0_488, WellA2_5_488, WellA2_6_488)
smoops_all_green$Total <- smoops_all_green$Area * smoops_all_green$Mean
smoops_all_yellow <-rbind(WellA1_561, WellA2_0_561, WellA2_5_561, WellA2_6_561)
smoops_all_yellow$Total <- smoops_all_yellow$Area * smoops_all_yellow$Mean
smoops_all_magenta <- rbind(WellA1_640, WellA2_0_640, WellA2_5_640, WellA2_6_640)
smoops_all_magenta$Total <- smoops_all_magenta$Area * smoops_all_magenta$Mean


ctrl_all_green <- rbind(WellA2_3_488, WellA2_4_488)
ctrl_all_green$Total <- ctrl_all_green$Area * ctrl_all_green$Mean

ctrl_all_yellow <- rbind(WellA2_3_561, WellA2_4_561)
ctrl_all_yellow$Total <- ctrl_all_yellow$Area * ctrl_all_yellow$Mean

ctrl_all_magenta <- rbind(WellA2_4_640)
ctrl_all_magenta$Total <- ctrl_all_magenta$Area * ctrl_all_magenta$Mean

```

## Normalising values to control of each color; to be able to compare values from different fluorophores used.

## First, calculating the average value of Total intensity for non-smOOPs in each fluorphore used.
```{r}
mean_ctrl_green <- mean(ctrl_all_green$Total)
mean_ctrl_yellow <- mean(ctrl_all_yellow$Total)
mean_ctrl_magenta <- mean(ctrl_all_magenta$Total)


smoops_all_green$Total_on_av <- smoops_all_green$Total/mean_ctrl_green
smoops_all_yellow$Total_on_av <- smoops_all_yellow$Total/mean_ctrl_yellow
smoops_all_magenta$Total_on_av <- smoops_all_magenta$Total/mean_ctrl_magenta

ctrl_all_green$Total_on_av <- ctrl_all_green$Total/mean_ctrl_green
ctrl_all_yellow$Total_on_av <- ctrl_all_yellow$Total/mean_ctrl_yellow
ctrl_all_magenta$Total_on_av <- ctrl_all_magenta$Total/mean_ctrl_magenta

```


## Merging the data in a single dataframe. 
```{r warning=FALSE}

# Calculate control means for each color
mean_ctrl_green <- mean(ctrl_all_green$Total, na.rm = TRUE)
mean_ctrl_yellow <- mean(ctrl_all_yellow$Total, na.rm = TRUE)
mean_ctrl_magenta <- mean(ctrl_all_magenta$Total, na.rm = TRUE)

# Define a function to process datasets correctly
process_well <- function(df, well_name, group_name, ctrl_mean) {
  df <- df %>%
    mutate(Total = Area * Mean) %>%  # Step 1: Calculate Total
    summarise(
      across(everything(), \(x) mean(x, na.rm = TRUE)),  # Step 2: Compute means
      Total_on_av = mean(Total, na.rm = TRUE) / ctrl_mean  # Step 3: Normalize Total
    ) %>%
    mutate(
      Well = well_name,
      Group = group_name
    )
  return(df)
}


# Compute summaries for each dataset
summary_list <- list(
  # Smoops (Green)
  process_well(WellA1_488, "WellA1_488", "smoops", mean_ctrl_green),
  process_well(WellA2_0_488, "WellA2_0_488", "smoops", mean_ctrl_green),
  process_well(WellA2_5_488, "WellA2_5_488", "smoops", mean_ctrl_green),
  process_well(WellA2_6_488, "WellA2_6_488", "smoops", mean_ctrl_green),
  
  # Smoops (Yellow)
  process_well(WellA1_561, "WellA1_561", "smoops", mean_ctrl_yellow),
  process_well(WellA2_0_561, "WellA2_0_561", "smoops", mean_ctrl_yellow),
  process_well(WellA2_5_561, "WellA2_5_561", "smoops", mean_ctrl_yellow),
  process_well(WellA2_6_561, "WellA2_6_561", "smoops", mean_ctrl_yellow),
  
  # Smoops (Magenta)
  process_well(WellA1_640, "WellA1_640", "smoops", mean_ctrl_magenta),
  process_well(WellA2_0_640, "WellA2_0_640", "smoops", mean_ctrl_magenta),
  process_well(WellA2_5_640, "WellA2_5_640", "smoops", mean_ctrl_magenta),
  process_well(WellA2_6_640, "WellA2_6_640", "smoops", mean_ctrl_magenta),
  
  # Control (Green)
  process_well(WellA2_3_488, "WellA2_3_488", "ctrl", mean_ctrl_green),
  process_well(WellA2_4_488, "WellA2_4_488", "ctrl", mean_ctrl_green),
  
  # Control (Yellow)
  process_well(WellA2_3_561, "WellA2_3_561", "ctrl", mean_ctrl_yellow),
  process_well(WellA2_4_561, "WellA2_4_561", "ctrl", mean_ctrl_yellow),
  
  # Control (Magenta)
  process_well(WellA2_4_640, "WellA2_4_640", "ctrl", mean_ctrl_magenta)
)



# Combine all summarized data into one dataframe
final_df <- bind_rows(summary_list)

# Remove the first two columns and replace them with 'Well' and 'Group'
final_df <- final_df %>%
  select(-c(1, 2)) %>%  # Remove the first two columns
  select('Well', 'Group', everything())  # Add 'Well' and 'Group' at the beginning


# Add gene name and gene_id (inserting the gene names of targets used in each well and fluorophore). 
# At the same time, added the values of expression (TPM) in semi-extractability library. 

final_df <- final_df %>%
  left_join(targets_info %>% select(Well, gene_id, gene_name, semi_naive), by = c("Well" = "Well"))

```

## Figure 1H 
## Plot the distribution of foci size (area in squared microns)

```{r}

t.test(final_df$Area ~ final_df$Group)

ggplot(final_df, aes(x = Group, y = Area, fill = Group)) +
  geom_boxplot() +
  geom_jitter() +
  scale_fill_manual(values = c("#bdcae2", "#4682B4")) + 
  showSignificance( c(1,2), 0.4, -0.02, "**") +
  labs(
    title = "Area size of foci for smOOPS (n=12) and non-smOOPs (n=5)",
    x = NULL,
    y = "Foci size (Area, um2) "
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15),
    title = element_text(size = 15)
  )

```


## To determine the fraction of the signal that is nuclear, I used another macro on all the images, which generated a nuclear mask (based on DAPI channel) and measured the intensity of foci within the nuclear mask and the whole frame of acquisition.

## By summing the foci intensity in nucleus and the whole frame, the nuclear fraction was calculated as the ratio between nuclear signal and the whole frame signal.

```{r include=FALSE, warning=FALSE}

# Define the folder path
folder_path <- "C:/Users/tajdak/UK Dementia Research Institute Dropbox/Tajda Klobucar_DRI/KI/Protocols/Semi_oops project/HCR_FISH design/IJS_imaging/july_2024/Particle_analysis"

# Define file pairs
file_pairs <- list(
  c("Nuc_well_A1_488.xlsx", "Frame_well_A1_488.xlsx"),
  c("Nuc_well_A2_0_488.xlsx", "Frame_well_A2_0_488.xlsx"),
  c("Nuc_well_A2_5_488.xlsx", "Frame_well_A2_5_488.xlsx"),
  c("Nuc_well_A2_6_488.xlsx", "Frame_well_A2_6_488.xlsx"),
  c("Nuc_well_A1_561.xlsx", "Frame_well_A1_561.xlsx"),
  c("Nuc_well_A2_0_561.xlsx", "Frame_well_A2_0_561.xlsx"),
  c("Nuc_well_A2_5_561.xlsx", "Frame_well_A2_5_561.xlsx"),
  c("Nuc_well_A2_6_561.xlsx", "Frame_well_A2_6_561.xlsx"),
  c("Nuc_well_A1_640.xlsx", "Frame_well_A1_640.xlsx"),
  c("Nuc_well_A2_0_640.xlsx", "Frame_well_A2_0_640.xlsx"),
  c("Nuc_well_A2_5_640.xlsx", "Frame_well_A2_5_640.xlsx"),
  c("Nuc_well_A2_6_640.xlsx", "Frame_well_A2_6_640.xlsx"),
  c("Nuc_well_A2_3_488.xlsx", "Frame_well_A2_3_488.xlsx"),
  c("Nuc_well_A2_4_488.xlsx", "Frame_well_A2_4_488.xlsx"),
  c("Nuc_well_A2_3_561.xlsx", "Frame_well_A2_3_561.xlsx"),
  c("Nuc_well_A2_4_561.xlsx", "Frame_well_A2_4_561.xlsx"),
  c("Nuc_well_A2_4_640.xlsx", "Frame_well_A2_4_640.xlsx")
)

# Initialize an empty data frame to store results
ratios_df <- data.frame(Well = character(), Ratio = numeric(), stringsAsFactors = FALSE)

# Load all Nucleus and Frame data into lists and calculate ratios
for (pair in file_pairs) {
  
  # Define file paths
  nuc_file <- file.path(folder_path, pair[1])
  frame_file <- file.path(folder_path, pair[2])

  # Read data from the files
  nuc_data <- read_excel(nuc_file)
  frame_data <- read_excel(frame_file)
  
  # Convert Area and Mean columns to numeric (in case they are read as characters)
  nuc_data$Area <- as.numeric(nuc_data$Area)
  nuc_data$Mean <- as.numeric(nuc_data$Mean)
  frame_data$Area <- as.numeric(frame_data$Area)
  frame_data$Mean <- as.numeric(frame_data$Mean)

  # Calculate the "Total" column for both files
  nuc_data$Total <- nuc_data$Area * nuc_data$Mean
  frame_data$Total <- frame_data$Area * frame_data$Mean

  # Sum "Total" values
  nuc_total <- sum(nuc_data$Total, na.rm = TRUE)
  frame_total <- sum(frame_data$Total, na.rm = TRUE)

  # Avoid division by zero
  if (frame_total == 0) {
    warning(paste("Frame total is zero for", frame_file, "- returning NA"))
    ratio_value <- NA
  } else {
    # Calculate the ratio
    ratio_value <- nuc_total / frame_total
  }

  # Use the full filename (nuc_file) as the well name
  well_name <- basename(nuc_file)  # Extract filename from the full path

  # Append the ratio and well name to the results data frame
  ratios_df <- rbind(ratios_df, data.frame(Well = well_name, Ratio_nuc = ratio_value))
}

# View the final data frame with Well names and Ratios
print(ratios_df)

# Adjust well names to match final_df
ratios_df$Well <- gsub("^Nuc_well_A", "WellA", ratios_df$Well)  # Replace 'Nuc_' with 'Well_'
ratios_df$Well <- gsub(".xlsx$", "", ratios_df$Well)  

# Append this to final_df dataframe

final_df <- final_df %>%
  left_join(ratios_df %>% select(Well, Ratio_nuc), by = c("Well" = "Well"))


```

## Figure S1F
## Plot the expression of all targeted transcripts (obtained from semi-extractability assay, TPM)

```{r}

t.test(final_df$semi_naive ~ final_df$Group)

ggplot(final_df, aes(x = Group, y = semi_naive, fill = Group)) +
  geom_boxplot() +
  geom_jitter() +
  scale_fill_manual(values = c("#bdcae2", "#4682B4")) + 
  labs(
    title = "Expression in semi library of smOOPS (n=12) and non-smOOPs (n=5)",
    x = NULL,
    y = "Expression (TPM)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15),
    title = element_text(size = 15)
  )

```


## Figure S1G 
## (left) Plot total intensity (log2 transformed) for all targeted transcripts
## (right) Plot the fraction of nuclear signal (on average for each targeted transcript)

```{r}
# Figure S1G (left)


t.test(final_df$Total_on_av ~ final_df$Group)

ggplot(final_df, aes(x = Group, y = Total_on_av, fill = Group)) +
  geom_boxplot() +
  geom_jitter() +
  scale_fill_manual(values = c("#bdcae2", "#4682B4")) + 
  scale_y_continuous(trans = "log2") +
  showSignificance( c(1,2), 7, -0.5, "*") +
  labs(
    title = "Total intensity of smOOPs (n=12) and non-smOOPs (n=5)",
    x = NULL,
    y = "log2(Total intensity)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15),
    title = element_text(size = 15)
  )

# Figure S1G (right)

ggplot(final_df, aes(x = Group, y = Ratio_nuc, fill = Group)) +
  geom_boxplot() +
  geom_jitter() +
  scale_fill_manual(values = c("#bdcae2", "#4682B4")) + 
  labs(
    title = "Nuclear fraction of foci smOOPS (n=12) and control (n=5)",
    x = NULL,
    y = "Fraction of nuclear signal"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 15),
    axis.title = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15),
    title = element_text(size = 15)
  )

```

